---
title: "The Beta Distribution"
author: "Matthew Stephens"
date: 2017-01-25
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```


# Overview

The purpose of this vignette is to introduce the Beta distribution. You should be familiar with basic concepts related to distributions before - e.g. maybe you have come across the normal distribution and a uniform distribution before, and understand what it would mean to talk about their mean, variance and density.

If you want more details you could look at [Wikipedia](https://en.wikipedia.org/wiki/Beta_distribution).

# The Beta Distribution

The Beta distribution is a distribution on the interval $[0,1]$. Probably you have come
across the $U[0,1]$ distribution before: the uniform distribution on $[0,1]$. 
You can think of the Beta distribution as a generalization of this that allows for some simple non-uniform distributions for values between 0 and 1. 

The Beta distribution has two parameters, which we will call $a$ and $b$. These two parameters determine the shape of the Beta distributions (just as the mean and variance determine the shape of the normal distribution). 

Following the usual convention, we will write $X \sim Be(a,b)$ as shorthand for "$X$ has a Beta distribution with parameters $a$ and $b$".

# Density

If $X \sim Be(a,b)$ then the density of $X$ is:
$$f_X(x) = \frac{1}{B(a,b)} x^{a-1}(1-x)^{b-1} \qquad (x \in [0,1]).$$

For those of you that are interested, $B(a,b)$ is known as the "beta function" and
is given by the integral 
$$B(a,b) = \int_0^1 x^{a-1} (1-x)^{b-1} \,dx.$$ 
This is where the beta distribution gets its name: its density involves the beta function.
However, for this introduction you do not have to worry very much about what $B(a,b)$ is: think of it as a constant (in that it does not depend on $x$), that is included so that the density integrates to 1, as all densities must. 

Because the Beta distribution is widely used, R has the built in function `dbeta` to compute this density. We will use this to look at some examples of the Beta distribution below.

# Examples

First we will look at some examples for $a=b$, with both $\geq 1$:
```{r}
p = seq(0,1, length=100)
plot(p, dbeta(p, 100, 100), ylab="density", type ="l", col=4)
lines(p, dbeta(p, 10, 10), type ="l", col=3)
lines(p, dbeta(p, 2, 2), col=2) 
lines(p, dbeta(p, 1, 1), col=1) 
legend(0.7,8, c("Be(100,100)","Be(10,10)","Be(2,2)", "Be(1,1)"),lty=c(1,1,1,1),col=c(4,3,2,1))
```


Now non-equal values of $a$ and $b$ with both $\geq 1$:
```{r}
p = seq(0,1, length=100)
plot(p, dbeta(p, 900, 100), ylab="density", type ="l", col=4)
lines(p, dbeta(p, 90, 10), type ="l", col=3)
lines(p, dbeta(p, 30, 70), col=2) 
lines(p, dbeta(p, 3, 7), col=1) 
legend(0.2,30, c("Be(900,100)","Be(90,10)","Be(30,70)", "Be(3,7)"),lty=c(1,1,1,1),col=c(4,3,2,1))
```


From these examples you should note the following:

  - The distribution is roughly centered on $a/(a+b)$. Actually, it turns out that the mean is exactly $a/(a+b)$. Thus the mean of the distribution is determined by the *relative* values of $a$ and $b$.
  - The larger the values of $a$ and $b$, the smaller the variance of the distribution about the mean. 
  - For moderately large values of $a$ and $b$ the distribution looks visually "kind of normal", although unlike the normal distribution the Beta distribution is restricted to [0,1].  
  - The special case $a=b=1$ is the uniform distribution.
  
## Values of $a,b <1$

The parameters $a$ and $b$ can also be less than 1, but the distribution in this case starts to have a different kind of shape. Specifically if $a<1$ then there is a peak at 0, and if $b<1$ then there is a peak at 1 (so if both are $<1$ then the distribution is U-shaped). Here are some examples:
```{r}
p = seq(0,1, length=100)
plot(p, dbeta(p, 0.1, 0.1), ylab="density", type ="l", col=4)
lines(p, dbeta(p, 0.5, 0.5), type ="l", col=3)
lines(p, dbeta(p, 0.1, 0.5), col=2) 
lines(p, dbeta(p, 0.5, 2), col=1) 
legend(0.5,2, c("Be(0.1,0.1)","Be(0.5,0.5)","Be(0.1,0.5)", "Be(0.5,2)"),lty=c(1,1,1,1),col=c(4,3,2,1))
```

## Exercise

- Sketch what you think the Be(5,5) and Be(0.5,5) and Be(500,200) distributions would look like. Check your sketches against the truth computed using `dbeta`. 

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
