---
title: "Simple example of how a p value translates to a Bayes Factor"
author: "Matthew Stephens"
date: 2016-04-20
bibliography: citations.bib
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

# Pre-requisites

You should know what a Bayes Factor is and what a p value is.

# Overview

Sellke et al [@Sellke.et.al.01] study the following question (paraphrased and shortened here). 

Consider the situation in which experimental drugs D1; D2; D3; etc are to be tested. Each test will be thought of as completely independent; we simply have a series of tests so that we can
explore the frequentist properties of p values. In each test,
the following hypotheses are to be tested:
$$H_0 : D_i \text{ has negligible effect}$$ versus
$$H_1 : D_i \text{ has a non-negligible effect}.$$ 

Suppose that one of these tests results in a p value $\approx p$. The question we consider is: How strong is the evidence that the drug in question has a non-negligible
effect?

The answer to this question has to depend on the distribution of effects under $H_1$.
However, Sellke et al derive a bound for the Bayes Factor. Specifically they show that,
provided $p<1/e$, the BF in favor of $H_1$ is not larger than 
$$1/B(p) = -[e p \log(p)]^{-1}.$$
(Note: the inverse comes from the fact that here we consider the BF in favor of $H_1$, whereas Sellke et al consider the BF in favor of H_0).

Here we illustrate this result using Bayes Theorem to do calculations under a simple scenario.

# Details

Let $\theta_i$ denote the effect of drug $D_i$. We will translate the null $H_0$ above to
mean $\theta_i=0$. We will also make an assumption that the effects of the
non-null drugs are normally distributed $N(0,\sigma^2)$, where the value of $\sigma$ 
determines how different the typical effect is from 0.

Thus we have:
$$H_{0i}: \theta_i = 0$$
$$H_{1i}: \theta_i \sim N(0,\sigma^2)$$.

In addition we will assume that we have data (e.g. the results of a drug trial)
that give us imperfect information about $\theta$. Specifically we assume $X_i | \theta_i \sim N(\theta_i,1)$. This implies that:
$$X_i | H_{0i} \sim N(0,1)$$
$$X_i | H_{1i} \sim N(0,1+\sigma^2)$$

Consequently the Bayes Factor (BF) comparing $H_1$ vs $H_0$ can be computed as follows:
```{r}
BF= function(x,s){dnorm(x,0,sqrt(s^2+1))/dnorm(x,0,1)}
```
Of course the BF depends both on the data $x$ and the choice of $\sigma$ (here `s` in the code).

We can plot this BF for $x=1.96$ (which is the value for which $p=0.05$):
```{r}
s = seq(0,10,length=100)
plot(s,BF(1.96,s),xlab="sigma",ylab="BF at p=0.05",type="l",ylim=c(0,4))
BFbound=function(p){1/(-exp(1)*p*log(p))}
abline(h=BFbound(0.05),col=2)
```
Here the horizontal line shows the bound on the Bayes Factor computed by Sellke et al.

And here is the same plot for $x=2.58$ ($p=0.01$):
```{r}
plot(s,BF(2.58,s),xlab="sigma",ylab="BF at p=0.01",type="l",ylim=c(0,10))
abline(h=BFbound(0.01),col=2)
```



Note some key features:

+ In both cases the BF is below the bound given by Sellke et al, as expected.
+ As $\sigma$ increases from 0 the BF is initially 1, rises to a maximum, and then gradually decays. This behavior, which occurs for any $x$, perhaps helps provide intuition into why it is even possible to derive an upper bound. 
+ In the limit as $\sigma \rightarrow \infty$ it is easy to show that (for any $x$) the BF $\rightarrow 0$. This is an example of "Bartlett's paradox", and illustrates why you should not just use a "very flat" prior for $\theta$ under $H_1$: the Bayes Factor will depend on how flat you mean by "very flat", and in the limit will always favor $H_0$. 

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```

## References
