---
title: "Every Bayesian computation is an integral (or a sum)"
author: "Matthew Stephens"
date: 2016-05-02
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

# Pre-requisites

Be familiar with basic probability and Bayesian calculations.

# Overview

The goal here is simply to point out that everything you want to compute in Bayesian calculations is an integral.

# Examples

Consider inference for a parameter $\theta$ from data $D$.

The posterior distribution of $\theta$ is 
given by Bayes Theorem

$$p(\theta | D) = p(\theta) p(D | \theta)/ p(D)$$

First note that the denominator $p(D)$ is an integral:
$$p(D) = \int p(D | \theta) p(\theta) d\theta$$.

Now suppose we want to estimate $\theta$ by its posterior mean.
This is
$$E(\theta | D) = \int \theta p(\theta |D ) d\theta.$$


And if we want to find a 90% posterior credible interval for $\theta$ then we want to find $A$ and $B$ such that $\Pr(\theta \in [A,B] | D) = 0.9$.
Note that the LHS of this is
$$\Pr(\theta \in [A,B]|D) = \int I(\theta \in [A,B]) p(\theta | D) d\theta,$$
where $I(E)$ denotes the indicator function for the event $E$, which takes the value 1 if $E$ is true and $0$ otherwise. 


# Examples: discrete

Of course, if $\theta$ is discrete then the integrals above all become sums.

For example
$$E(\theta | D) = \sum_n \theta_n \Pr(\theta=\theta_n | D)$$
where $\theta_1,\theta_2,\dots$ are the possible values for $\theta$.

# Summary

Pretty much all the things you want to compute when doing Bayesian inference are integrals (or sums) of one kind or another...

If you are computing 1-dimensional integrals then numerical methods
are often useful. For example, Simpsons Rule, Gaussian Quadrature.
These can also work in 2-dimensions, and maybe even 3 or 4.

Other simple methods that can work for low dimensions: naive Monte Carlo, and Importance Sampling. Also Laplace approximation.

For higher dimensions we usually resort to Markov Chain Monte Carlo.



## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
