---
title: "Untitled"
author: "First Last"
date: YYYY-MM-DD
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

# Pre-requisites

Be familiar with the [likelihood ratio and log-likelihood ratio for discrete data](likelihood_ratio_simple_models.html).


# Combining independent data

One nice feature of LLRs is that when you combine independent data, then LLRs add. 

To be more precise, suppose you observe two independent pieces of data $x_1$ and $x_2$.
Then 
$$LLR(M_1,M_0; x_1,x_2) = LLR(M_1,M_0; x_1) + LLR(M_1,M_0; x_2).$$

This follows directly from the definition of LR, and the fact that for independent pieces
of data the likelihoods multiply (so log-likelihoods add).

For example, in our tusk example, the markers are considered independent, so the data at six markers
can be broken down into six independent pieces of data, one from each marker. Thus the LLR for
the data from 6 markers can be written as the sum of LLRs for each marker.
```{r}
x = c(1,0,1,0,0,1)
fS = c(0.40, 0.12,0.21,0.12,0.02,0.32)
fF = c(0.8,0.2,0.11,0.17,0.23,0.25)
L = function(f,x){ log(f^x*(1-f)^(1-x)) }
LR = L(fS,x)/L(fF,x)
print(LR)
```


# Definition

# Examples


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
