---
title: "Log-likelihood and log-likelihood ratio"
author: "Matthew Stephens"
date: 2016-01-05
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

# Pre-requisites

You should have come across the concept of [likelihood](likelihood.html), and be aware that likelihood provides
a measure of the *relative* support in the data for different values of a parameter.

# Overview

Our purpose here is simply to define and motivate the use of the log-likelihood and the log-likelihood ratio. 

For many reasons it is often convenient to work with the logarithm of the likelihood function rather that the likelihood function itself. This is referred to as the log-likelihood function, and usually denoted with $l$ (a lower-case $L$):
$$l(\theta):=\log_e L(\theta).$$

Similarly, it is often convenient to work with the log-likelihood-ratio, rather than the likelihood ratio.
Recall that the likelihood ratio (LR) for $\theta=\theta_1$ vs $\theta=\theta_0$ is given by 
$$LR(\theta_1,\theta_0) = L(\theta_1)/L(\theta_0).$$
Therefore the logarithm of this, 
known as the log-likelihood-ratio, is 
$$\log_e(LR(\theta_1,\theta_0)) = l(\theta_1)-l(\theta_0).$$

Recall that the likelihood function $L(\theta)$ provides a measure of the *relative* support in the data for different values of the parameter $\theta$. In particular, when comparing the likelihood for two different values of $\theta$, what matters is the likelihood ratio. Similarly, when comparing the log-likelihood for different values of $\theta$, what matters is the difference $l(\theta_1)-l(\theta_0)$ (the log-likelihood ratio).

# Examples

We can plot the log-likelihood-ratio for the coin-tossing example as follows.
```{r}
theta = seq(0,1,length=100)
l =function(theta){2*log(theta) + log(1-theta)}
plot(theta, l(theta)-l(0.5), xlab="theta", ylab="l(theta)-l(0.5)",type="l")
```







## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
