---
title: "Time reversibility of a Markov Chain"
author: "John Novembre"
date: 2016-01-31
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

This is a work-in-progress.

```{r}
knitr::opts_chunk$set(eval = FALSE)
```

# Pre-requisites

TBD

# Overview

TBD

# Time reversibility of a Markov Chain

TBD

# Example: Gary's mood

In Sheldon Ross's Introduction to Probability Models, he has an example (4.3) of a Markov Chain for modeling Gary's mood.  Gary alternates between 3 state: Cheery ($X=1$), So-So ($X=2$), or Glum ($X=3$).  Here we input the $P$ matrix given by Ross and we input an arbitrary initial probability matrix.

```{r}
# Define prob transition matrix 
# (note matrix() takes vectors in column form so there is a transpose here to switch col's to row's)
P=t(matrix(c(c(0.5,0.4,0.1),c(0.3,0.4,0.3),c(0.2,0.3,0.5)),nrow=3))
# Check sum across = 1
apply(P,1,sum)  

# Definte initial probability vector
x0=c(0.1,0.2,0.7)
# Check sums to 1
sum(x0)
```



#### Time-reversibility

The reverse chain transition probabilities are: $$\frac{\pi_j P_{ji}}{\pi_i}$$
```{r}
diag(1/pivec)%*%(t(P)%*%diag(pivec))
```
So for our example we see it's not time reversible...

Let's make a time-reversible example...
```{r}
P=t(matrix(c(c(0.5,0.25,0.25),c(0.25,0.45,0.3),c(0.25,0.3,0.45)),nrow=3))
# Check sum across = 1
apply(P,1,sum)  

# And the stationary distribution would be
pi0=(P%^%100)[1,]
# Now we check time reversibility
diag(1/pi0)%*%t(P)%*%diag(pi0)
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
