---
title: "Linear combinations of independent normals are normal"
author: "Matthew Stephens"
date: "2022-03-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


# Pre-requisites

Basic familiarity with the univariate normal distribution

# A statement of the basic property

The simple goal of this vignette is to introduce 
a basic property of the (univariate) normal distribution: that linear combinations of independent normal variables are also normal.

Formally, suppose $Z_1$ and $Z_2$ represent independent, normally distributed random variables. Then for any scalars $a$ and $b$, the linear combination 
$$X:=aZ_1+bZ_2$$ 
is also (univariate) normal.

Also, by basic properties of expectation and variance, $E(X) = aE(Z_1)+bE(Z_2)$ and $var(X)= a^2 var(Z_1) + b^2 var(Z_2)$.



# Example

The following code provides a visual illustration of this idea with $a=2$ and $b=3$,
but it holds for any $a$ and $b$.

First we sample some values of $X$ by randomly 
generating $Z_1$ and $Z_2$, and computing $X=aZ_1+bZ_2$:
```{r}
Z1 = rnorm(1000)
Z2 = rnorm(1000)
a = 2
b = 3

X = a*Z1 + b*Z2 
```

The property says that the samples of $X$ look normal. A quick histogram and qqplot suggest it does.... (of course this is not a proof that the property holds; it is just an illustration of the idea).
```{r}
hist(X)
qqnorm(X)
```


# Addendum: Stable distributions

If you are curious by nature, you might now ask:
is the normal distribution the only distribution that satisfies this property?  
The answer is "no". For example, $t$ distributions also satisfy this property.
Distributions that satisfy this property are called  ``stable" distributions.
You can read more at [https://en.wikipedia.org/wiki/Stable_distribution]

