---
title: "Ornstein-Uhlenbeck Process"
author: "Yanyu Liang"
date: 2018-03-08
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

# Pre-requisites

Please read [Brownian motion](brownian_motion_and_guassian_process.html) beforehand.

# Motivation

Consider the drunk person sernario again but now instead of walking along a flat land, the guy is walking a convex one. Although he still tends to walk in random direction but the curvature of the line traps him/her because of gravity (keeping the person not far from the minimum point, $x_0$). Suppose the trapping force gets bigger proportionally to the distance to $x_0$. Namely, for an time interval $t_i \rightarrow t_i + \Delta t = t_{i + 1}$, $\Delta_{\text{force}} x = -\alpha (x(t_i) - x_0) = -\alpha (x_i - x_0)$, where $\alpha$ controls how strong the force can be. Then, we can simulate the process as follow.

```{r}
trappedRandomWalk <- function(nsteps, initial.pos, alpha, x0, sigma = 1) {
  pos <- rep(NA, nsteps + 1)  # generate a vector to record the positions as time goes
  pos[1] <- initial.pos  # set initial position
  for(i in 2 : (nsteps + 1)) {  # loop over time points
    delta.force <- - alpha * (pos[i - 1] - x0)  # compute how force affects the update
    delta.random <- rnorm(1, mean = 0, sd = sigma)  # compute how random walk affects the update
    delta.x <- delta.force + delta.random
    pos[i] <- pos[i - 1] + delta.x
  }
  return(pos)
}
```

```{r, echo=FALSE}
simulationWrapper <- function(nrepeat, sim.name, nsteps, initial.pos, alpha, x0) {
  sim <- lapply(1 : nrepeat, trappedRandomWalk, nsteps = nsteps, initial.pos = initial.pos, alpha = alpha, x0 = x0)
  sim <- data.frame(sim)
  colnames(sim) <- paste0(sim.name, 1 : nrepeat)
  sim$time <- 1 : nrow(sim)
  sim <- melt(sim, id.vars = c('time'))
  sim$alpha <- alpha
  return(sim)
}
library(ggplot2)
library(reshape2)
library(stringr)
set.seed(2018)
nrepeat <- 10
# perform the simulation 
# with parameters as follow
# x0 = 0, alpha = 0.1, sigma = 1, initial.pos = 100
x0 <- 0
initial.pos <- 200
nsteps <- 200

sim0 <- simulationWrapper(nrepeat = nrepeat, sim.name = 'sim0', nsteps = nsteps, initial.pos = initial.pos, alpha = 1, x0 = x0)

sim1 <- simulationWrapper(nrepeat = nrepeat, sim.name = 'sim1', nsteps = nsteps, initial.pos = initial.pos, alpha = 0.1, x0 = x0)

# simulate under alpha = 0.01
sim2 <- simulationWrapper(nrepeat = nrepeat, sim.name = 'sim2', nsteps = nsteps, initial.pos = initial.pos, alpha = 0.01, x0 = x0)

# compare with a pure random walk (w/o) trapping (alpha = 0)
sim3<- simulationWrapper(nrepeat = nrepeat, sim.name = 'sim3', nsteps = nsteps, initial.pos = initial.pos, alpha = 0, x0 = x0)
```

Perform the simulation under $\alpha = 0.1, 0.01, 0$ with $x_0 = `r x0`, \sigma = 1, x(0) = `r initial.pos`$. Note that when $\alpha = 0$, the process is equivalent to random walk since force has not contribution. Red line is the average $x_i$.

```{r, echo=FALSE}
# plot trajectories
df.plot <- rbind(sim0, sim1, sim2, sim3)
ggplot(df.plot, aes(x = time, y = value, group = variable)) + geom_line() + stat_summary(fun.y = mean, geom = "line", lwd = .5, aes(group = 1), color = 'red')+ facet_grid(alpha ~ ., labeller = label_both) + ylab('X(time)')
```

As you can see, when the force is big, the person goes to $x_0$ fairly quickly and if $\alpha$ is close to zero, the process behaves more like random walk. Note that even when the drunk person has already close to $x_0$, the variance of the walk afterward is still smaller than pure random walk. This is because the random effect is not completely cummulated as in random walk case. In other word, any random effect is dumped by force and the bigger the force the smaller the variance would be.

# 

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
