---
title: "Example of difficulty of calibrating p values"
author: "Matthew Stephens"
date: 2017-04-17
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

# Pre-requisite 

You should know what a $p$ value is.


# Background

A key problem with $p$ values, when testing null hypotheses, is that they
can be difficult to calibrate. That is, it is hard to answer the question
"If I get a $p$-value of 0.01 (or any other number) how strong is the evidence against the null hypothesis?"

## Example

Here we just give a simple (but artificial) example of an test in which a $p$ value
of 0.01 actually corresponds to evidence *for* the null, even though 0.01 is usually
considered to be strong evidence *against* the null. 
(This example is modified from the book Bayesian Analysis, by J Berger, p25.) 

Suppose $x \in \{1,2,3\}$ and $\theta \in \{0,1\}$
with 

```{r table, echo=FALSE}
df = rbind(c("$p(x | \\theta=0)$", 0.005,0.005,0.99),c("$p(x | \\theta=1)$", 0.999,0.001,0))
knitr::kable(df,col.names=c("x","1","2","3"))
```

Note that the [likelihood ratios](likelihood_ratio_simple_models.html) 
for $H_1$ vs $H_0$ for $x=1,2,3$ are $999/5, 1/5$ and $0$ respectively. So as $x$ increases the evidence against $H_0$ decreases.


Now, let us suppose that we observe $x=2$.
Then by definition the $p$ value for this observation is
$$p:= \Pr(\text{we would see evidence as strong or stronger against $H_0$ than $x=2$} | \theta=0).$$

Here "evidence as strong or stronger against $H_0$ than $x=2$" is $x \in \{1,2\}$.  And the probability of this under $H_0$ is 
$$\Pr(x \in \{1,2\} | H_0) = 0.005+0.005 = 0.01.$$

So the $p$ value for $x=2$ is 0.01.

And yet, the observation $x=2$ is 5 times more probable under $H_0$ than under $H_1$! So
$x=2$ has $p$ value 0.01 but is actually evidence *for* $H_0$.


## Discussion

This example is obviously contrived to make a point: so it only demonstrates
that it is possible to contrive a situation
where $p=0.01$ corresponds to evidence *for* $H_0$. 

However, given this it seems natural to ask: in "typical" situations, does $p=0.01$ correspond
to evidence for or against $H_0$? Of course, the answer to this depends on what
one views as "typical". For a start towards answering this question see 
[here](BF_and_pvalue.html).






## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
