---
title: Gibbs sampler Exercises for a simple genetic mixture (with answers)
author: "Matthew Stephens and Eric C Anderson"
date: "2021-07-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
 
This vignette contains a set of exercises to lead you through creating a [Gibbs sampler](gibbs1.html) for a [simple genetic mixture](r_simplemix.html). 
It relies on code from the latter vignette, including the code
to complete the exercises. 
This first chunk contains all the code you need 
and is taken from [here](r_simplemix_withanswers.html). 

```{r}
# Original code to sample haploid genotypes

#' @param n number of samples
#' @param p a vector of allele frequencies
r_haploid_genotypes = function(n,p){
  R = length(p)
  x = matrix(nrow = n, ncol=R)
  for(i in 1:n){
    x[i,] = rbinom(R,rep(1,R),p)
  }
  return(x)
}

# Exercise 1 answerto sample from a simple genetic mixture model

#' @param n number of samples
#' @param P a 2 by R matrix of allele frequencies
#' @param w a w-vector of component prior probabilities
r_simplemix = function(n,P,w=c(0.5,0.5)){
  R = ncol(P) # number of loci
  z = rep(0,n) # used to store population of origin of each individual
  x = matrix(nrow = n, ncol=R) #used to store genotypes
  
  for(i in 1:n){
    z[i] = sample(1:2, size=1, prob=w)
    x[i,] = r_haploid_genotypes(1,P[z[i],])
  }
  return(list(x=x,z=z))
}


### Exercise 2 answer

#' Compute the likelihood for allele frequences p given genotype data x
#' @param p an R vector of allele frequencies
#' @param x an R vector of genotypes
likelihood = function(p,x){
  return(prod(p^x*(1-p)^(1-x)))
}

#' normalize a vectorx, so it's elements sum to 1
#' @param x a vector
normalize = function(x){x/sum(x)}

#' Compute posterior assignment probabilities
#' @param x an n by R matrix of genotypes
#' @param P a 2 by R matrix of allele frequencies
#' @param w a 2-vector of prior probabilities
posterior_prob_assignment=function(x,P,w){
  n = nrow(x)
  K = length(w)
  post = matrix(nrow=n, ncol=K) # to store the posterior probabilities
  lik = rep(0,K) # a vector to store the likelihoods
  for(i in 1:n){
    for(k in 1:K){
      lik[k] = likelihood(P[k,],x[i,])
    }
    post[i,] = normalize(lik*w)
  }
  return(post)
}


### Exercise 3 answer

#' Compute the posterior parameters for allele frequencies given genotypes 
#' and population labels
#' @param x an n by R matrix of genotypes
#' @param z an n vector of population assignments
#' @param a a 2-vector of prior parameters
posterior_param_allele_frequencies=function(x,z,a){
 K = 2
 R = ncol(x)
 post_param = array(dim=c(2,K,R))
 for(k in 1:K){
   for(r in 1:R){
     number_of_ones = sum(x[z==k,r]) #number of 1s observed at locus r in population k
     total = sum(z==k) # how many individuals in population k
     post_param[,k,r] = a + c(number_of_ones, total - number_of_ones)
   }
 }
 return(post_param)
}
```


## Exercises

For this exercise we will assume the mixture proportions `w` are fixed and known. The things we do not know are the assignment indicators (`Z`)
and the population allele frequences (`P`).
The Gibbs sampler for fitting the mixture model must therefore alternative between two key steps: sample the assignment indicators given the allele frequencies,
and sample the allele frequencies given the indicators. 
That is:

- sample from `Z | P, x, w`
- sample from `P | Z, x, w`


The following three exercises build up this Gibbs sampler.
Each exercise includes an answer template (you need to fill in the code in `...`) and a small chunk of code you can use to try out your answer.

Note: all these chunks have the option set `` so that this document compiles even though the code is incomplete. You will have to remove these if you want to run these chunks on compilation.

## Exercise 1: `newZ_gibbs()`

Write a function `newZ_gibbs <- function(x, P, w) {...}` to sample from
the full conditional distribution for the population assignment indicators `Z`. It should use `x` , `P`, and `w` and the function `posterior_prob_assignment()` given above to compute this full conditional probabiities for the assignments, and then, using that, sample a new value of `z_i` for every individual.

Here is an answer template:
```{r }
#' Simulate new values for the population assignments from their full conditional distribution
#' @param x an n by R matrix of genotypes
#' @param P a 2 by R matrix of allele frequencies
#' @param w a 2-vector of prior probabilities
newZ_gibbs <- function(x, P, w) {
    K <- nrow(P)
    n <- nrow(x)
    
    # get the full conditionals
    zprobs <- posterior_prob_assignment(x, P, w)  # makes a matrix of probabilities
    
    newZ <- rep(NA, n)
    
    for (i in 1:n) {
      newZ[i] <- sample(1:K, size = 1, prob = zprobs[i,])   # simulate new Z's from the zprobs
    }
    
    newZ
}
```


When you have filled this in try running it on example data:
```{r }
# simulate example data
p = seq(0.1,0.9,length=9)
P = rbind(p,1-p)
set.seed(123)
sim = r_simplemix(n=20,P)

newZ <- newZ_gibbs(sim$x, P, w = c(0.5, 0.5))

# here is the original value
sim$z
# here is the new value
newZ
```

## Exercise 2. `newP_gibbs()`

Write a function `newP_gibbs <- function(x, z, a) {...}` to
sample from
the full conditional distribution for the population allele frequencies `P`. 
It should use `x`, `z`, and `a` and the function `posterior_param_allele_frequencies()` to find the parameters of the beta distributions that are the full conditional distributions for the allele frequencies, and then sample new values of the allele frequencies, giving a new matrix `P`.


Here is an answer template:
```{r}
#' Simulate new values for the allele frequencies 
#' @param x an n by R matrix of genotypes
#' @param z an n vector of population assignments
#' @param a a 2-vector of prior parameters
newP_gibbs <- function(x,z,a = c(1,1)) {
  
  # compute the beta parameters for the full conditionals of P
  p_params <- posterior_param_allele_frequencies(x, z, a)
  
  R <- ncol(x)  # number of loci
  K <- 2

  # prepare a matrix for output
  newP <- matrix(nrow = 2, ncol = R)
  
  # simulate beta random variables to fill the P matrix
  for(r in 1:R) {
    for(k in 1:K) {
    
    # get the beta parameters for population k
    par1 <- p_params[1,k,r]
    par2 <- p_params[2,k,r]
    
    # simulate a new value from a beta distribution
    newP[k,r] <- rbeta(1, par1, par2)
    }
  }
  
  newP
}
```


Again, when you have completed it, try running on example data:
```{r}
set.seed(123)
sim = r_simplemix(n=20,P)
newP <- newP_gibbs(sim$x, sim$z, a = c(1,1))

# here is the original value
P
# here is the new value
newP
```



## Exercise 3: put it together

Write a function `mcmc_gibbs_Z_and_P <- function(
  x, 
  a = c(1, 1), 
  w=c(0.5, 0.5), 
  nIter = 500
)` implementing a Gibbs sampler to fit the simple genetic mixture model.
Start from a random initial value of `Z`, then alternately update `P` and `Z` using `newZ_gibbs()` and `newP_gibbs()` functions.


Here is an answer template:

```{r }
# do mcmc by Gibbs sampling for Z and P
#' @param x an n by R matrix of genotypes
#' @param a a 2-vector of prior parameters for the allele frequencies
#' @param w a 2-vector of mixing proportions assumed to be (0.5, 0.5)
#' @param nIter the number of sweeps (updates of both Z and P) of the chain to perform
mcmc_gibbs_Z_and_P <- function(
  x, 
  a = c(1, 1), 
  w=c(0.5, 0.5), 
  nIter = 500
) {
  
  n <- nrow(x)
  R <- ncol(x)
  K <- 2
  
  # we need to have an initial value for the Z's.  Just sample it, randomly from w
  z <- sample(1:2, size = n, replace = TRUE, prob = w)
  
  # cycle over the sweeps
  for (rep in 1:nIter) {
    
    # Gibbs update for P
    P <- newP_gibbs(x, z, a)
    
    # Gibbs update for z
    z <- newZ_gibbs(x, P, w)
    
    # For now, just print the current values
    # print z
    print(paste("Z: rep =", rep))
    print(z)
    
    # print P
    print(paste("P: rep =", rep))
    print(P)
  }
}
```



Here we can run this on the example data:
```{r}
set.seed(123)
sim = r_simplemix(n=20,P)

# now, given only the genotypes x, the allele frequency prior parameters a,
# and assuming mixing proportions of w = c(0.5, 0.5), we can run a Markov chain
# that samples from the posterior distribution of Z, and P.

# Print the true values of z and P from the simulated data:
sim$z
P

# Currently, the function just prints the values of Z and P each iteration.
# Just run it for 10 iterations
mcmc_gibbs_Z_and_P(sim$x, nIter = 10)
```

If you go back and check, you should find that the initial `Z` does not look anything like the true value from the simulation, but, after only 10 iterations, `Z` in the Markov chain is quite
similar to the true `Z`.

