---
title: 'simple example: bayesian inference for binomial proportion'
runtime: shiny
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

The code chunks below cannot be rendered as part of the website. [GitHub Pages](https://pages.github.com) only serves static webpages, whereas shiny applications are dynamic.

```{r eval-false}
library("knitr")
opts_chunk$set(eval = FALSE)
```

This R Markdown document is made interactive using Shiny. Unlike the more traditional workflow of creating static reports, you can now create documents that allow your readers to change the assumptions underlying your analysis and see the results immediately. 

To learn more, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change.  This demonstrates how a standard R plot can be made interactive by wrapping it in the Shiny `renderPlot` function. The `selectInput` and `sliderInput` functions create the input widgets used to drive the plot.

```{r, echo=TRUE}
inputPanel(
#  selectInput("n_breaks", label = "Number of bins:",
#              choices = c(10, 20, 35, 50), selected = 20),
  
    sliderInput("prior_alpha", label = "alpha for prior",
              min = 0, max = 1000, value = 1, step = 0.2),
   
     sliderInput("prior_beta", label = "beta for prior",
              min = 0, max = 1000, value = 1, step = 0.2)
)

renderPlot({
  q=seq(0,1,length=100)
  plot(q,dbeta(q,as.numeric(input$prior_alpha),as.numeric(input$prior_beta)),type="l", xlab= "allele frequency, q")
 # dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  #lines(dens, col = "blue")
})
```




## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
