---
title: "Untitled"
author: "First Last"
date: YYYY-MM-DD
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

# Pre-requisites

include the most complex concepts required to understand the material.

# Overview

Suppose we have a logistic regression $Y_i | X_i \sim Bern(p_i)$
where 
$$log(p_i/(1-p_i)) = \mu + \theta X_i.$$

We will assume that $X_i \in {-1,+1}$, and assume
priors for $\mu$ and $\theta$: 
$$\mu \sim N(0,100)$$
$$\theta \sim N(0,1)$$

For illustration we simulate data where $\mu=\theta=0$:
```{r}
x = sample(c(-1,1),1000,replace=TRUE)
y = rbinom(1000,1,0.5)

#b is a vector b=(mu,theta)
#loglikelihood for logistic regression
loglik = function(b){
  eta = b[1]+b[2]*x
  p = exp(eta)/(1+exp(eta))
  return(sum(log(y*p+(1-y)*(1-p))))
}

#b is a vector b=(mu,theta)
log_prior = function(b){
  return(dnorm(b[1],0,10, log=TRUE)+dnorm(b[2],0,1,log=TRUE))
}

#b is a vector b=(mu,theta)
log_post = function(b){
  return(loglik(b)+log_prior(b))
}
```

Let's compute a 95% CI for $\theta$.
First try a discrete grid

**Note:** This is still a work in progress.

```{r eval=FALSE}
m = seq(-10,10,length=100)
t = seq(-2,2,length=100)
df = expand.grid(m=m,t=t)
head(df)
#df = c(df,dplyr::ddply(df,log_post))

```

# Definition

# Examples


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
