---
title: "Use PCA to visualize single-cell data"
author: "Joyce Hsiao"
date: 2015-12-12
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```

## Objectives

We import data collected in [singleCellSeq][] and visualize the data using principal component analysis. The PCA code was created by [John Blischak][JB] and are stored in *ashlar/codes*. 

[singleCellSeq]: http://jdblischak.github.io/singleCellSeq/analysis
[JB]: http://jdblischak.github.io

PCA visualization was performed to inspect the first two major sources of variation. Do cells come from different individuals share similar variation? Do cells come from different batches share similar variation? 


## Prepare data

Import gene by sample (cell) matrix. Each column consists of molecule counts of a single cell. 

```{r}
molecule_final <- read.table("../data/molecules-final.txt",
                             header = TRUE, 
                             sep = "\t",
                             stringsAsFactors = FALSE,
                             row.names = 1)
```

Import metadata information. This annotates for every cell the individual (cell line) labels, replicate labels, etc. 

```{r}
annotation_filter <- read.table("../data/annotation-filter.txt",
                                header = TRUE,
                                sep = "\t",
                                stringsAsFactors = FALSE)
```


## Making PCA plots

Source PCA functions.

```{r}
source("../code/functions.R")
```

Run PCA.

```{r}
pca_results <- run_pca(as.matrix(molecule_final))
```

Visualize PCA results.

```{r}
plot_pca(pca_results$PCs,
         explained = pca_results$explained,
         metadata = annotation_filter,
         color = "individual",
         shape = "batch") +
  labs(title = "Molecules (log2 cpm) for single cells")
```



## Session information

```{r info}
sessionInfo()
```
