---
title: "Discrete Markov Chains: Finding the Stationary Distribution"
author: "Matt Bonakdarpour; with edits by Matthew Stephens and Jitao David Zhang"
date: 2016-01-21
---




# Pre-requisites

This document assumes basic familiarity with Markov chains and linear algebra.

# Overview

In this note, we illustrate one way of analytically obtaining the stationary distribution for a finite discrete Markov chain.

# 3x3 example

Assume our probability transition matrix is:
$$P = \begin{bmatrix}
    0.7 & 0.2 & 0.1 \\
    0.4 & 0.6 & 0 \\
    0   & 1   & 0 
\end{bmatrix}$$

Since every state is *accessible* from every other state, this Markov chain is irreducible. Every irreducible finite state space Markov chain has a unique stationary distribution. Recall that the stationary distribution $\pi$ is the row vector such that
$$\pi = \pi P$$. 

Therefore, we can find our stationary distribution by solving the following linear system:
$$\begin{align*}
0.7\pi_1 + 0.4\pi_2  &= \pi_1 \\
0.2\pi_1 + 0.6\pi_2 + \pi_3 &= \pi_2 \\
0.1\pi_1 &= \pi_3
\end{align*}$$
subject to $\pi_1 + \pi_2 + \pi_3 = 1$. Putting these four equations together and moving all of the variables to the left hand side, we get the following linear system:
$$\begin{align*}
-0.3\pi_1 + 0.4\pi_2  &= 0 \\
0.2\pi_1 + -0.4\pi_2 + \pi_3 &= 0 \\
0.1\pi_1 - \pi_3 &= 0 \\
\pi_1 + \pi_2 + \pi_3 &= 1
\end{align*}$$

We will define the linear system in matrix notation:
$$\underbrace{\begin{bmatrix}
    -0.3 &  0.4 & 0  \\
     0.2 & -0.4 & 1  \\
     0.1 &    0 & -1  \\
     1   &  1   & 1 
\end{bmatrix}}_A \begin{bmatrix}
\pi_1 \\
\pi_2 \\
\pi_3 
\end{bmatrix} = \underbrace{\begin{bmatrix}
0 \\
0 \\ 
0 \\
1
\end{bmatrix}}_b \\
A\pi^T = b$$

The stationary distribution, which is usually represented by a row vector, is
transposed with $\pi^T$.

Since this linear system has more equations than unknowns, it is an overdetermined system. Overdetermined systems can be solved using a QR decomposition, so we use that here. (In brief, `qr.solve` 
works by finding the QR decomposition of $A$, $A=QR$ with $Q'Q=I$ and $R$ an upper triangular matrix. Then if $A\pi^T = b$ it must be the case that $QR\pi^T=b$ which implies $R\pi^T = Q'b$, and this can be solved easily because $R$ is triangular.)


```{r}
A        <- matrix(c(-0.3, 0.2, 0.1, 1, 0.4, -0.4, 0, 1, 0, 1, -1, 1 ), ncol=3,nrow=4)
b        <- c(0,0,0, 1)


```{r}
pi        <- qr.solve(A,b)
names(pi) <- c('state.1', 'state.2', 'state.3')
pi 
```

We find that:
$$\begin{align*}
\pi_1 \approx 0.54, \pi_2 \approx 0.41, \pi_3 \approx 0.05
\end{align*}$$

Therefore, under proper conditions, we expect the Markov chain to spend more time in states 1 and 2 as the chain evolves.


## The General Approach
Recall that we are attempting to find a solution to 
$$\pi = \pi P$$
such that $\sum_i \pi_i =1$. First we rearrange the expression above to get:
$$\begin{align}
\pi - \pi P &= 0 \\
\pi (I - P) &= 0 \\
(I - P)^T\pi^T &= 0
\end{align}$$

One challenge though is that we need the constrained solution which respects that $\pi$ describes a probability distribution (i.e. $\sum \pi_i = 1$).  Luckily this is a linear constraint that is easily represented by adding another equation to the system.  So as a small trick, we need to add a row of all 1's to our $(I-P)^T$ (call this new matrix $A$) and a 1 to the last element of the zero vector on the right hand side (call this new vector $b$). Now we want to solve $A\pi = b$ which is over-determined so we solve it as above using `qr.solve`.

The function `stationary` below implements the general approach, and we test it
with the worked example above.

```{r}
stationary <- function(transition) {
  stopifnot(is.matrix(transition) &&
	    nrow(transition)==ncol(transition) &&
            all(transition>=0 & transition<=1))
  p <- diag(nrow(transition)) - transition
  A <- rbind(t(p),
	     rep(1, ncol(transition)))
  b <- c(rep(0, nrow(transition)),
         1)
  res <- qr.solve(A, b)
  names(res) <- paste0("state.", 1:nrow(transition))
  return(res)
}
stationary(matrix(c(0.7, 0.2, 0.1, 0.4, 0.6, 0, 0, 1, 0),
                  nrow=3, byrow=TRUE))
```

